---
title: "Re-evaluating Growth in the Time of Debt"
subtitle: "[DSLC stages]: Data cleaning and pre-processing"
format: 
  gfm: default
execute:
  echo: true
editor: source
number-sections: true
editor_options: 
  chunk_output_type: console
---

## Domain problem formulation


## Data source overview



## Step 1: Review background information {#sec-bg-info}


### Data dictionary

`debt.xls` is a wide data set which has the following variables:
Each row refers to its own country, each column refers to its own year, and the data collected is the percentage public sector debt relative to Gross Domestic Product (GDP). In other words, what is the government's debt relative to the country's economy size?

| Variable                                 | Description                                                       |
| :--------------------------------------- | :---------------------------------------------------------------- |
| Central Government Debt (Percent of GDP) | The country name of interest                                      |
| 1950 ... 2024                            | The year of the estimate (Central Government Debt Percent of GDP) |
| Estimates start after                    | The year before these estimates were created                      |

The data is sourced from the International Monetary Fund, which calculates this value by estimating the Gross Domestic Product (GDP) and public debt of a country, then performing (debt / gdp) * 100. It is important that by this definition, we would expect this value to range from zero to infinity (note that public debt can be larger than GDP.)

`growth.csv` is another wide data set which has the following variables

| Variable        | Description                                   |
| :-------------- | :-------------------------------------------- |
| County.Name     | The full name of country                      |
| Country.Code    | A three letter code for a country             |
| Indicator.Name  | The parameter of interest (always GDP growth) |
| Indicator.Code  | The code for the parameter of interest        |
| X1960 ... X2025 | The year of the estimate                      |

The data is sourced from the World Bank Ground (WBG), which collects its data from official country statistics, Organisation for Economic Co-operation and Development (OECD), and other sources.

## Step 2: Loading in the data

### Debt Data

```{r load debt data, message = FALSE, warning = FALSE}
library(readxl)

debt_raw <- read_xls("../data/debt.xls")

dim(debt_raw)

class(debt_raw$`1950`)
```

This mostly makes sense. Data from this set goes from 1950 to 2024 (74 + 3 columns).

We should consider removing/renaming unnecessary/confusingly named columns.
We may also want to replace "no data" with `NA` such that our entry values can be numeric.

```{r numeric debt data, message = FALSE, warning = FALSE}
library(tidyverse)

debt_raw_2 <- debt_raw |> mutate(
  across(
    -1,

    ~ as.numeric(ifelse(.x == "no data", NA, .x))
  )
)
```

### Annual GDP Growth Data

```{r load gdp data}
gdp_raw <- read.csv("../data/growth.csv")

dim(gdp_raw)

summary(gdp_raw$X1960)
summary(gdp_raw$X1961)
```

From a cursory glance, this data set has fewer entries than the debt data set. It will be important to see which countries appear in both. It is also important to note that data does not exist for 1960 (which may be set as a baseline.)

## Step 3: Examine the data and create action items

### Debt Data

#### Finding invalid values
We should make our columns numeric when possible. Therefore we need to replace "no data" entries with `NA`.
This will be achieved using the following function.

```{r debt invalid values}
source("functions/data_cleaning.R")

debt <- replace_non_numeric(debt_raw, "no data")
```

The first two rows are empty, so we can ignore them.

```{r debt invalid rows}
debt <- debt[-c(1, 2), ]
```

Rows 239 onward do not represent countries. We can remove them.

```{r debt unneeded rows}
debt <- debt[1:238, ]
```

#### Assessing Column names

The first column should be renamed to better reflect how it represents a country.

```{r debt rename countries}
debt <- debt |> rename(country_name = `Central Government Debt (Percent of GDP)`)
```

`Estimates start after` contains no useful information, we will remove this column.

```{r debt remove final column}
debt <- debt |> select(-`Estimates start after`)
```

In the end, it may be helpful to create a long table. We will create a new column called `debt_pct_gdp` for this.

```{r debt pivot long}
debt_long <- debt |> pivot_longer(
  cols = -1,
  names_to = "year",
  values_to = "debt_pct_gdp"
)
```

#### Assessing variable type

```{r debt variable type}
class(debt_long$country_name)

debt_long <- debt_long |> mutate(year = as.numeric(year))
class(debt_long$year)

class(debt_long$debt_pct_gdp)
```

#### Assessing data completeness

Are there any missing combinations?

```{r debt completeness}
expected <- expand_grid(
  country_name = unique(debt_long$country_name),
  year = unique(debt_long$year)
)

missing <- expected |>
  anti_join(debt_long, by = c("country_name", "year"))

nrow(missing)
```

We can see that each combination is present. However it is important to note that we do not always have data for each country-year combination.

```{r debt summary}
summary(debt["2024"])
```

We are missing data for many countries in each year (with 2024 as an example). This can be for a few reasons (for historical data, the country may not have existed, and other countries may not have stable governments/reliable reporting of these figures.) This is something that must be noted.

### GDP Data

#### Finding invalid values

This data does have appropriate `NA` values for missing data.

An issue with this data set is that there is no easy way to separate countries and non-countries. This must be recognized when performing a join between our two data sets.

#### Assessing column names

`Country.Name` should be rewritten as `county_name` in accordance to snake case. Each year column is prefixed by an `X`. This may be easier to remove once columns have been pivoted longer. The column generated by `values_to` for this pivot should be named `growth_pct_gdp`. Other columns will be unnecessary.

#### Assessing variable type

`country_name` should be a character variable, `year` should be a numeric (or date), and `growth_pct_gdp` should be a numeric.


## Step 4: Clean and pre-process the data

Let's recreate the tables shown for Australia from 1960 to 1969.

```{r two table output}
clean_debt_data(debt_raw) |> filter(country_name == "Australia", year %in% 1960:1969)

clean_growth_data(gdp_raw) |> filter(country_name == "Australia", year %in% 1960:1969)
```

To join the two data sets, we need to look at the country names from each output.

```{r country names}
debt_country_names <- clean_debt_data(debt_raw)$country_name |> unique()
growth_country_names <- clean_growth_data(gdp_raw)$country_name |> unique()

setdiff(debt_country_names, growth_country_names)
setdiff(growth_country_names, debt_country_names)
```

The debt data set has more countries, but all countries in the GDP growth data set are contained in the debt data set. We can now continue with the join.

```{r data join}
join_country_data_sets(debt_raw, gdp_raw) |> filter(country_name == "Australia", year %in% 1960:1969)
```

We now have a data set to do analysis with. $\square$